An examination of the accessibility implications of a pilot COVID-19 vaccination program in Hamilton, Ontario

### 2021-05-04 Accessibility-Vaccination-Sites-Hamilton/Response-to-Reviewers-v2.Rmd

• I find it more interpretable and impactful when you all present travel time results measured in minutes rather than in person-hours. I understand needing the person-hours results as a cumulative, individual-based measure to compare across income and region classes. However, as a reader, I do find the person-hours analysis to be harder to understand than the travel time analysis. I suggest the authors put some more thought towards explaining, particularly, the person-hours analysis results clearly.

>\textcolor{blue}{We agree. The cumulative values are easier to understand in hours, but at the individual level it is less intuitive to say "0.09 hours". We have adjusted this in the revised paper.}

• I like the idea of the comparison scenario, but wonder how the scenario sites were chosen. Could the three sites feasibly serve as vaccination sites? Were these locations “optimized” in any way to lead to greater reductions in travel time disparities? Were they randomly chosen? I think more is needed to clarify this scenario experiment/analysis for the reader.

>\textcolor{blue}{The three sites we use for the scenario are existing pharmacies and of chains used by the province elsewhere for the pilot, so they are definitely feasible for the program. We did not use any systematic approach to select these sites, we just went with three reasonable sites. The fact that three sites chosen in a reasonable but not systematic way can lead to such important reductions is telling about the effort that the province put in the selection of sites for the program. There are certainly better ways of approaching this issue (for instance, by designing optimal accessibility landscapes, see Paez et al., 2013) but we feel that designing optimal landscapes is beyond the scope of a short findings paper, since the finding that we wish to report is the inequity in the roll-out of the program.}

@@

\newpage

# Reviewer 2

This paper looks at a location allocation issue that has been studied before in geograohy for several decades, where to place new facilities ti reduce travel and increase access to heakthcare in a region. The study finds that rural and low income are poorly srved and big burden of travel falls on these groups and they offer the location of new sites for vaccinations to decrease such burden. The paper depnds ona major assumotion that those who commuted by mode X will commute by the same mode for vaccination, which is not true, if COVID has shiwn us something is new travel behaviour emerge and peopel are switching modes so this assumption need to be highlighted as a major limitation. it could have been better to see a map with the old sites alone, with new sites added and have the background as the collective travel time this will enable the reader to visualize the impact spatially.

>>\textcolor{blue}{Thank you for these comments. Both authors are familiar with the literature on accessibility to healthcare, having contributed to it as recently as 2019. The findings that we wish to report here concern the specific situation of the vaccination pilot in Ontario. This is a policy-relevant issue for which there is no empirical information available besides this paper. We agree that a limitation of the analysis is that we assume all demographic/travel behavior patterns are stable since 2016. This assumption was already highlighted in the original paper (and aren't all assumptions limitations by definition?); in the revision we  redouble our effort to highlight this issue more explicitly as a limitation. We are aware that travel behavior for routine activities has changed during the pandemic (the first author has researched this in, for example, Paez, 2020, and Jamal and Paez, 2020, both papers published in Findings). But travel to get an essential vaccine is not necessarily a routine, quotidian sort of travel behavior. We do not really know (at least for the region under study) how behavior for non-routine travel may have changed during the pandemic since we know relatively little about non-routine travel to begin with. In principle, it seems possible to do scenario analysis, but for this we would need to make up numbers regarding modal shares during the pandemic, and such an endeavor is not without problems either - besides probably requiring a substantially longer paper. We plotted the locations of the pharmacies in the travel time maps for ease of visualization.}

### 2021-05-04 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.Rmd
minor typographic edits

Along with the provision of health care facilities to treat severe cases of COVID-19 [e.g., @Ghorbanzadeh2021spatial; @Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs. The Government of Ontario, in Canada, implemented in 2021 a pilot program to offer vaccines in pharmacies for people 60 years and older. In the first incarnation of this program, the province approved 325 pharmacies in Toronto, Windsor-Essex, and Kingston (the latter including the region of Frontenac-Lennox and Addington). The City of Hamilton, the third most populous urban center in the province, operated dedicated vaccination centers for people aged 70+ and mobile pop-up clinics for people aged 75+; however, it did not have any approved pharmacies under the pilot for other populations until April 1st 2021, with the announcement of 20 pharmacies in the city \footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}. At the same time that the expansion of the pilot was announced, the province extended eligibility for this program to people 55 years and older.

@@

We concentrate on the 55 to 69 years old population segment because the older 70+ group have access to other dedicated facilities besides those in the provincial pilot (including pop-up clinics for people 75+).

@@

The top panel of Figure \ref{fig:maps-baseline} shows the average multi-modal travel times (weighted by modal shares) by TAZ in Hamilton. It is apparent that travel times tend to be lower in much of suburban Hamilton and higher in the urban core and some rural parts of the city, particularly to the west. This is unsurprising, given the higher probability of travel by car and the predominantly suburban character of the vaccination sites. However, even accounting for the distribution of population, this leads to large disparities in the number of person-hours of travel across the city, with a concentration of the burden of travel in the urban core and the rural west (see bottom panel of Figure \ref{fig:maps-baseline}).

@@

As seen in Table \ref{tab:distribution-results}, under the pilot program approximately `r round((table_comparison[5,2] + table_comparison[4,2]) / (sum(table_comparison[1:5,2])) * 100, 2)`\% of people live in DAs in the bottom 40% of the median household income scale, but they account for `r round((table_comparison[5,3] + table_comparison[4,3]) / (sum(table_comparison[1:5,3])) * 100, 2)`\% of the total person-hours of travel. In contrast, `r round((table_comparison[1,2] + table_comparison[2,2]) / (sum(table_comparison[1:5,2])) * 100, 2)`\% of people aged 55 to 69 in DAs in the top 40% of the median household income scale accrue only `r round((table_comparison[1,3] + table_comparison[2,3]) / (sum(table_comparison[1:5,3])) * 100, 2)`\% of the total person-hours of travel. Where the mean travel time of residents of DAs with high median household income is `r table_comparison[1, 4] %>% set_units(h) %>% set_units(min) %>% round() %>% drop_units()` minutes, residents of lower income DAs average `r table_comparison[5, 4] %>% set_units(h) %>% set_units(min) %>% round() %>% drop_units()` minutes in travel time. In addition to longer average travel time, residents in lower income DAs also see substantially larger variations in travel times, and some may face considerably longer travel times (see top-left panel in Figure \ref{fig:results}).

@@

For comparison purposes we consider a scenario with some modest additions to the list of pharmacies in the provincial pilot. For the scenario we choose three plausible sites (i.e., pharmacies that belong to chains already used by the province for this program). We do not conduct a systematic search of candidate locations. We note that some papers propound the design optimal accessibility landscapes [e.g., @Horner2008optimal; @Paez2013exploring], but such an endeavor is beyond the scope of a short findings paper. Instead, we repeat the analysis after including the three urban sites shown in white circles in Figure \ref{fig:pharmacies-and-regions}.

@@

The results indicate that the locations chosen by the province for the pilot vaccination program do not serve urban or rural residents of the city well, and there are some important questions regarding equity of access to the program. As shown, a disproportionate burden in the cost of travel (in terms of travel time) fall on lower income urban populations and rural populations. Selection of three sensible urban locations does much to alleviate disparities in the burden of transportation - and the fact that this is so despite the lack of a systematic search for optimal locations does not speak well of the effort put by the province in the choice of vaccination sites. The situation is different for rural populations, mainly because there are not many candidate locations in rural parts of the city; in this case, it is likely that increasing access for residents there will necessitate an expansion of the existing mobile vaccination pop-up clinic program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.

>>>>

Along with the provision of health care facilities to treat severe cases of COVID-19 [e.g., @Ghorbanzadeh2021spatial; @Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs. The Government of Ontario, in Canada, implemented a pilot program to offer vaccines in pharmacies for people 60 years and older in early 2021. In the first incarnation of this program, the province approved 325 pharmacies in Toronto, Windsor-Essex, and Kingston (the latter including the region of Frontenac-Lennox and Addington). The City of Hamilton, the third most populous urban center in the province, operated dedicated vaccination centers for people aged 70+ and mobile pop-up clinics for people aged 75+; however, it did not have any approved pharmacies under the pilot for other populations until April 1st 2021, with the announcement of 20 pharmacies in the city \footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}. At the same time that the expansion of the pilot was announced, the province extended eligibility for this program to people 55 years and older.

@@

We concentrate on the 55 to 69 year old population segment because the older 70+ group have access to other dedicated facilities besides those in the provincial pharmacy pilot (including pop-up clinics for people 75+).

@@

The top panel of Figure \ref{fig:maps-baseline} shows the average multimodal travel times (weighted by modal shares) by TAZ in Hamilton. It is apparent that travel times tend to be lower in much of suburban Hamilton and higher in the urban core and some rural parts of the city, particularly to the west. This is unsurprising, given the higher probability of travel by car and the predominantly suburban character of the vaccination sites. However, even accounting for the distribution of population, this leads to large disparities in the number of person-hours of travel across the city, with a concentration of the burden of travel in the urban core and the rural west (see bottom panel of Figure \ref{fig:maps-baseline}).

@@

As seen in Table \ref{tab:distribution-results}, under the pilot program approximately `r round((table_comparison[5,2] + table_comparison[4,2]) / (sum(table_comparison[1:5,2])) * 100, 2)`\% of people live in DAs in the bottom 40% of the median household income scale, but they account for `r round((table_comparison[5,3] + table_comparison[4,3]) / (sum(table_comparison[1:5,3])) * 100, 2)`\% of the total person-hours of travel. In contrast, `r round((table_comparison[1,2] + table_comparison[2,2]) / (sum(table_comparison[1:5,2])) * 100, 2)`\% of people aged 55 to 69 in DAs in the top 40% of the median household income scale accrue only `r round((table_comparison[1,3] + table_comparison[2,3]) / (sum(table_comparison[1:5,3])) * 100, 2)`\% of the total person-hours of travel. Where the mean travel time of residents of DAs with high median household income is `r table_comparison[1, 4] %>% set_units(h) %>% set_units(min) %>% round() %>% drop_units()` minutes, residents of lower income DAs average `r table_comparison[5, 4] %>% set_units(h) %>% set_units(min) %>% round() %>% drop_units()` minutes in travel time. In addition to greater average travel times, the top-left panel in Figure \ref{fig:results} suggests residents in lower income DAs also see substantially larger variability in travel times and some may face considerably longer trip durations.

@@

For comparison purposes we consider a scenario with some modest additions to the list of pharmacies in the provincial pilot. For the scenario we choose three plausible sites (i.e., pharmacies that belong to chains already used by the province for this program). We do not conduct a systematic search of candidate locations. We note that some papers propound the design of optimal accessibility landscapes [e.g., @Horner2008optimal; @Paez2013exploring], but such an endeavor is beyond the scope of a short findings paper. Instead, we repeat the analysis after including the three urban sites shown in white circles in Figure \ref{fig:pharmacies-and-regions}.

@@

The results indicate that the locations chosen by the province for the pilot vaccination program do not serve urban or rural residents of the city well, and there are some important questions regarding equity of access to the program. As shown, a disproportionate burden in the cost of travel (in terms of travel time) falls on lower income urban populations and rural populations. The selection of three sensible urban locations does much to alleviate disparities in the burden of transportation - and the fact that this is so despite the lack of a systematic search for optimal locations does not speak well of the effort put forth by the province in the choice of vaccination sites. The situation is different for rural populations, mainly because there are not many candidate locations in rural parts of the city; in this case, it is likely that increasing access for residents there will necessitate an expansion of the existing mobile vaccination pop-up clinic program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.
